Staged Abstract Interpreters

### 2019-09-04 paper/40_absinterp.tex
revision

\paragraph{Primitive Operations} The primitive operations are changed to adapt
the new monad stack scheme and value domains. A notable change is that the store
update operation joins the value with the existing values, for the same address.
As what mentioned before, we may have to explore both branches for conditionals:
in @br0@ method, we combine the results using @mplus@ from @MonadPlus@, which
requires the value domains are join-semilattices (in our case @Set@s and
@Map@s).
\begin{lstlisting}
  def set_store(αv: (Addr, Value)): AnsM[Unit] = liftM(StateTMonad.mod(σ => σ ⊔ Map(αv)))
  def br0(test: Value, thn: => Ans, els: => Ans): Ans = ReaderTMonadPlus.mplus(thn, els)
\end{lstlisting}

@@

For function applications, @ap_clo@ looks the same as the concrete interpreter.
The difference is now the first argument @rator@ is a set of closures (remember
that @Value@ is now an alias of @Set[AbsValue]@), so we can not directly
destruct it by pattern matching. Instead, we use an auxiliary function @lift_nd@
that takes a set and lifts the elements into the monad stack. Then we can still
implement @ap_clo@ in monadic style, where the closures come from the @SetT@ and
thus the nondeterminism can be naturally handled. The light gray code shows what
is added from the concrete @ap_clo@.
\begin{lstlisting}[escapechar=!]
  def lift_nd[T](vs: Set[T]): AnsM[T]
  def ap_clo(ev: Expr => Ans)(rator: Value, rand: Value): Ans = for {
    !\hl{CloV(Lam(x, e), $\rho$) <- lift\_nd[AbsValue](rator)}!
    α <- alloc(x)
    _ <- set_store(α → rator)
    rt <- local_env(ev(e))(ρ + (x → α))
  } yield rt
\end{lstlisting}

@@

As we mentioned earlier, the ADI approach uses a two-cache mechanism to
compute the least fixed-point and prevent non-termination, which is also called a
\textit{co-inductive} caching algorithm or \textit{truncated depth-first
evaluation} \cite{Rosendahl:AbsIntPL}. It has been used in other abstract
interpreter or fixed point computation
\cite{DBLP:journals/pacmpl/DaraisLNH17, Wei:2018:RAA:3243631.3236800,
  Rosendahl:AbsIntPL}. The idea is to use a @in@ cache and a @out@ cache during
the depth-first evaluation, where the @in@ cache stores the result from last
iteration, and @out@ cache is used for updates in current iteration. In the next
iteration, the last @out@ cache will be served as @in@ cache, and an empty cache
is plugged-in to the @out@ slot. Once the @out@ cache does not contain anything
new than the @in@ cache, the fixed-point is reached.

In our monad stack, the @in@ and @out@ cache modeled by a reader effect and
state effect, respectively. We first defined several methods to manipulate
the two caches through the monad stack; the implementations are elided.
\begin{lstlisting}
  def ask_in_cache: AnsM[Cache]; def get_out_cache: AnsM[Cache]
  def put_out_cache(out: R[Cache]): AnsM[Unit]
  def set_out_cache(cfg: R[Config], vs: R[(Value, Store)]): AnsM[Unit]
\end{lstlisting}

@@

The iteration terminates when the resulting @out@ cache is equivalent to the
input @in@ cache, which indicates that there is no more fact can be gained, thus
the iteration should end and we have reached the least fixed point.

@@

Finally, we override the top-level @run@ method by running the monadic value
with the empty abstract environment $\rho_0$, initial abstract store $\sigma_0$,
and initial caches @cache@$_0$.

>>>>

\paragraph{Primitive Operations} The primitive operations are changed according to
the new monad stack scheme and value domains. A notable change is that the store
update operation joins the new value with existing values, for the same address.
As we mentioned before, we may have to explore both branches for conditionals:
in @br0@ method, we combine the results using @mplus@ from @MonadPlus@, which
requires the value domains are join-semilattices (in our case @Set@s and
@Map@s).
\begin{lstlisting}
  def set_store(αv: (Addr, Value)): AnsM[Unit] = liftM(StateTMonad.mod(σ => σ ⊔ Map(αv)))
  def br0(test: Value, thn: => Ans, els: => Ans): Ans = ReaderTMonadPlus.mplus(thn, els)
\end{lstlisting}

@@

For function applications, @ap_clo@ looks the same as the concrete interpreter.
The difference is that now the first argument @rator@ is a set of closures (remember
that @Value@ is an alias of @Set[AbsValue]@), so we can not directly
destruct it by pattern matching. Instead, we use an auxiliary function @lift_nd@
that takes a set and lifts the elements into the monad stack. Then we can still
implement @ap_clo@ in monadic style, where the closures come from the @SetT@ and
thus the nondeterminism can be naturally handled. The light gray code shows what
is added from the concrete @ap_clo@.
\begin{lstlisting}[escapechar=!]
  def lift_nd[T](vs: Set[T]): AnsM[T]
  def ap_clo(ev: Expr => Ans)(rator: Value, rand: Value): Ans = for {
    !\hl{CloV(Lam(x, e), $\rho$) <- lift\_nd[AbsValue](rator)}!
    α <- alloc(x)
    _ <- set_store(α → rator)
    rt <- local_env(ev(e))(ρ + (x → α))
  } yield rt
\end{lstlisting}

@@

As we mentioned earlier, the ADI approach uses a two-cache mechanism to
compute the least fixed-point and prevent non-termination, which is also called a
\textit{co-inductive} caching algorithm or \textit{truncated depth-first
evaluation} \cite{Rosendahl:AbsIntPL}. It has been used in other abstract
interpreter or fixed point computation
\cite{DBLP:journals/pacmpl/DaraisLNH17, Wei:2018:RAA:3243631.3236800,
  Rosendahl:AbsIntPL}. The idea is to use a @in@ cache and a @out@ cache during
the depth-first evaluation, where the @in@ cache stores the result from last
iteration, and @out@ cache is used for updating in current iteration. In the next
iteration, the last @out@ cache will be served as @in@ cache, and an empty cache
is plugged-in to the @out@ slot. Once the @out@ cache does not contain anything
new than the @in@ cache, the fixed-point is reached.

In our monad stack, the @in@ and @out@ cache are modeled by a reader effect and
state effect, respectively. We first define several methods to manipulate
the two caches through the monad stack (the implementations are elided).
\begin{lstlisting}
  def ask_in_cache: AnsM[Cache]; def get_out_cache: AnsM[Cache]
  def put_out_cache(out: R[Cache]): AnsM[Unit]
  def set_out_cache(cfg: R[Config], vs: R[(Value, Store)]): AnsM[Unit]
\end{lstlisting}

@@

The iteration terminates when the resulting @out@ cache is equivalent to the
input @in@ cache, which indicates that there is no more fact can be discovered,
therefore the iteration should end and we have reached the least fixed point.

@@

Finally, we override the top-level @run@ method by running the monadic value
with an empty abstract environment $\rho_0$, initial abstract store $\sigma_0$,
and initial caches @cache@$_0$.
